Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-11 doc/response.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{natbib}
\setcitestyle{numbers} 
\setcitestyle{square}
\bibliographystyle{prsb}

@@

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit.
We apologize for the delay.
Reviewer 1 suggested that we should divide the paper into two: theory and application.
Given the relevance of this manuscript to the current COVID-19 outbreak, we think it is best to write an application-based one first (this one) and then write a theory-based on later.
We believe that incorporating some theory with application will help readers understand the material better at the moment.
Otherwise, we have tried our best to incorporate the reviewers' suggestions.
See below for detailed response.

\rev{Reviewer \#1}

\revtext{The main objective of this paper is to correct selection bias in observational
(contact tracing) data to yield unbiased estimate of the "true" distribution
corresponding to generation intervals defined in Lines 42-43 (if I understand
properly).}

The main objective is to estimate the effective generation-interval distribution, which implicitly accounts for spatial structure. We tried to make this clear now.

\revtext{Lines 42-43: "A generation interval is defined as the time between when a
person becomes infected and when that person infects another person."
(with reference [10]). Indeed, most of the cited literature in this paper
start with the definition exactly the same way. I feel that this definition
is imprecise and is not as obvious as it reads, especially in its relationship
with the intrinsic generation interval (Line 46) and its distribution (Lines
49,101). It is not proven that they are equivalent.}

We agree that this definition may be imprecise but we feel that it may be clearer for the readers to begin with a familiar definition. 
Instead, we now point to imprecision of this definition.
We also make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval) and distinguish it from the population-level distributions:

``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
while this definition is widely used in the literature, it is not connected directly to a population-level distribution.''

\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it
equivalent to the definition in Lines 42-43?}

We now define intrinsic generation interval properly:

``\emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population...''

\revtext{Line 46: "......intrinsic generation intervals measure the infectiousness of an
infected individual." This statement is vague. I will get back to this phrase
a few pages later in the discussion on "equilibrium distributions" and
suggest ways to make it more precise.}

\revtext{Lines 49,101: "...the intrinsic generation interval distribution..." While the
interval itself is not defined in Line 46 (in my opinion), its distribution is
clarified in Line 101 in relation to equation (2) under homogeneous mixing.
I hereby pose the following questions:}

\revtext{
1. Although $g(\tau) > 0$ satisfies all the properties of a probability density
function (p.d.f.) of a non-negative continues random variable, does that
automatically correspond to a specific random variable relevant to any of
the observable (or conceptually observable) "generation intervals"?
}

We now define $g(\tau)$ as follows: ``Expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population''

\revtext{
2. Is it the p.d.f. corresponding to the definition in Lines 42-43? Proof?
}

Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. We have tried to make this clearer in the text.

\revtext{
Lines 44-48: "... the observed generation-interval distribution can change...
...observed generation intervals refer to the time between actual infection
events." The actual infection events correspond to the exact moments of
passing the infection between a pair of infected-susceptible individuals.
Are they observable? In most cases, including contact tracing, observable
events are clinical (e.g. onset of symptoms). The relationship between
the interval deÖned by the time between actual infection events and the
interval deÖned by the time between clinical onsets is unclear. At least the
latter depends on the incubation period (of the infectee), in addition to the
latent period and the infectious period (of the infector). The incubation
period is defined as the time from infection to symptom onset, and is
different from the latent period such as in the SEIR model. This paper
does not make such distinction. It assumes that the time between actual
infection events "were observable" in a hypothetical way.
}

\revtext{
Suggestion: Re-phrase "observed generation intervals" or make some further
clarification on this. This is important because naive readers may tend
to take the face value and collect different kinds of observed data and
make inferences, such as that based on clinical events. This is potentially
misleading.
}

We agree that calling it ``observed'' can be misleading. We now refer to them as \emph{realized} intervals.

``There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.''

\revtext{Lines 54-55: ``the backward generation interval...'' This is a key measurement in this paper along with the length-bias in relation to data observation scheme. Conceptually, there may also be a "forward" generation
interval. Both concepts were initially raised in Svensson (2007, ref.[21]
in the paper), who pointed out the difference in distribution between the
primary generation time as measured prospectively from the time of the
infection of the infector to the transmission to the infectee, and the secondary generation time as measured retrospectively from the time of the
transmission to the infectee to the infection of the infector.}

We now introduce forward.

\revtext{1. does that mean that your objective is to make unbiased inference on the
"forward", or the primary, generation time?}

Thank you for pointing this out. This is what we want. We have tried to clarify this throughout the manuscript. We no longer use the term ``effective'' generation-interval distribution.

\revtext{2. does the definition in Lines 42-43 correspond to the forward generation
interval?}

No.

\revtext{3. does the definition in Lines 42-43 correspond to the intrinsic generation
interval (Line 46)?}

No.

\revtext{4. does the definition in Lines 42-43 correspond to a random variable with
p.d.f. given by $g(\tau)$?}

No.

\revtext{5. do you imply that the intrinsic generation interval (not defined in Line 46
but distributed according to Line 101) and the forward measure, and the
objective of the unbiased inference, are all equivalent?}

No.

\bibliography{networkprsb}
\end{document}

### 2020-05-11 doc/network.tex
edits

A generation interval is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
Due to individual variation in infection time, the observed generation-interval distribution can change depending on when and how it is measured \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.
There are important distinctions to be made when estimating generation intervals: \emph{intrinsic} generation intervals measure the infectiousness of an infected individual,
while \emph{observed} generation intervals refer to the time between actual infection events.
%% Observed generation intervals in turn can be measured either \emph{forward} in time by looking at a cohort of individuals infected at the same time and asking when they infected others or \emph{backward} in time by looking at a cohort and asking when their infectors were infected \citep{champredon2015intrinsic}.
% When aggregated generation intervals are observed until a given time in an ongoing epidemic (often via contact tracing), we refer to these as \emph{censored} (or right-censored) intervals.
% Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution \citep{champredon2015intrinsic}.
% As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles, 
% and the distribution of forward generation intervals will have shorter mean than the intrinsic distribution.

While the intrinsic generation-interval distribution is often assumed to be fixed, the shape of the observed generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}. 
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals (which look at a cohort of individuals, and ask when their infectors were infected)  will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.

In practice, generation intervals are often measured by aggregating available information from contact tracing. 
%% Whereas backward generation intervals represent infections that are one generation apart from a cohort of infectees, these aggregated generation intervals represent infections across multiple cohorts (i.e., the time of infection of the infectees can range anywhere from the beginning of an epidemic to the most recent infection).
While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
%% In fact, censored observations can be thought of as an average of backward generation-interval distributions, over the cohorts which have been infected before the observation time (and weighted by cohort size).
%% JD: We say this later anyway.

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
%% even in the early phase of an epidemic. 
This means that contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already) 
As a result, realized generation intervals will have shorter mean than the intrinsic generation-interval distribution.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.
These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
%% In other words, Trapman \textit{et al.}'s finding can be explained by the gap between the intrinsic generation interval and an ``effective'' (spatially corrected) generation interval.

>>>>

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level distribution.
There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case,
while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.

While the shape of intrinsic generation-interval distribution is often assumed to be invariant across time, the shape of the realized generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
Forward generation intervals can contract over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.

Forward generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, forward generation intervals will have shorter mean than the intrinsic generation-interval distribution.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In practice, realized generation intervals are often measured by aggregating available information from contact tracing.
While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
We further predict the effective generation-interval distribution to be equivalent to the \emph{initial} forward generation-interval distribution, which describes the renewal process of infection during the exponential growth rate.
